---
title: ''
author: "CJ"
date: "Tuesday, October 14, 2014"
output: html_document
---

### Loading and preprocessing the data   
First we load the data and remove the rows with NA
```{r}  
unzip("activity.zip");
data = read.table("activity.csv", sep=",", header=TRUE);  
data$date = as.character(data$date);  
data = data[complete.cases(data),];  
```  


### What is mean total number of steps taken per day?

To figure out the mean number of steps, let's first set a frame with the number of steps per day:  

```{r}
library(plyr)
sum.data <- ddply(data, c("date"), summarise,
               total.steps = sum(steps));
```

Now we can calculate the mean and median and do a histogram of the mean number of steps per day:  
```{r}
mean.steps = mean(sum.data$total.steps);
median.steps = median(sum.data$total.steps);
hist(sum.data$total.steps, main="Histogram total number of number of daily steps", xlab="daily steps", ylab="Frequency");
```

**The mean number of steps per day is `r mean.steps` and the median is `r median.steps` **



### What is the average daily activity pattern?

Let's figure it out:
```{r}
avg.steps.per.interval <- ddply(data, c("interval"), summarise,
               mean.steps = mean(steps));
with(avg.steps.per.interval, plot(interval, mean.steps, type = "l", main="Avg number of steps in 5-minute interval", xlab="5-minute interval", ylab="Number of steps"));
```
 

### Imputing missing values

For this part of the exercise, we must deal with the NA values, so let's start by re-reading the data:  

```{r}
data = read.table("./activity.csv", sep=",", header=TRUE);
data$date = as.character(data$date)
```

How many missing steps do we have?
```{r}
na.steps = data[which(is.na(data$steps)),];
#number of nas:
missing.steps.num = nrow(na.steps);
```
We have **`r missing.steps.num`** rows with missing steps.

Now, let's fill the missing steps with reasonable values:  
First calculate the average number of steps per day:
```{r}
avg.steps.per.day <- ddply(data, c("date"), summarise, .drop = FALSE, mean.steps = mean(steps));
```

Some days are missing all steps information, so replace it with the total average number of steps
```{r}
avg.steps.per.day$mean.steps=ifelse(is.na(avg.steps.per.day$mean.steps),mean.steps, avg.steps.per.day$mean.steps);
```
... and merge the average number of steps per day with the rest of the data:  
```{r}
data = merge(data, avg.steps.per.day, by.x="date", by.y="date");
```

For those intervals that have NA, replace the number of steps with the average on that day.
```{r}
num.of.intervals.in.a.day = 288;
data$date = as.Date(data$date, format = "%Y-%m-%d");
data$steps = ifelse(is.na(data$steps), data$mean.steps/num.of.intervals.in.a.day, data$steps);
```

How does the mean and the median of the filled-NA data compare with the mean and median of this data without the NA rows?
```{r}
sum.data <- ddply(data, c("date"), summarise,
               total.steps    = sum(steps));
new.mean.steps = mean(sum.data$total.steps);
new.median.steps = median(sum.data$total.steps);
hist(sum.data$total.steps, main="Total number of number of daily steps- After imputing the missing values", xlab="daily steps", ylab="Frequency");
```

After imputing the missing values, the mean number of steps per day is `r new.mean.steps` and the median is `r new.median.steps`.  These numbers are very close to those before imputing the missing values.


### Compare week days activity and weekend days activity:

Add a factor for weekdays and weekends:
```{r}
data$day.type <- as.factor(ifelse(weekdays(data$date) %in% c("Saturday", "Sunday" ) ,"weekend", "weekday"));
data.weekdays = data[which(data$day.type=="weekday"),];
data.weekends = data[which(data$day.type=="weekend"),];
avg.steps.per.interval.on.weekdays <- ddply(data.weekdays, c("interval"), summarise,
               mean.steps = mean(steps));
avg.steps.per.interval.on.weekends <- ddply(data.weekends, c("interval"), summarise,
               mean.steps = mean(steps));
par(mfrow = c(2, 1));
with(avg.steps.per.interval.on.weekdays, plot(interval, mean.steps, type = "l", main="Weekdays", ylab="Avg number of steps"));
with(avg.steps.per.interval.on.weekends, plot(interval, mean.steps, type = "l", main="Weekends", ylab="Avg number of steps"));
```
